The Impact of Bias in Landcover Maps

### 2015-09-22 paper/main.tex
Updated through Results with AW comments.

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
\usepackage{placeins}

@@

\title{Quantifying the impacts of bias in landcover data on global change analyses}

\author{Lyndon Estes\affil{1}{Princeton University, Princeton, NJ USA},
Peng Chen\affil{2}{Indiana University, Bloomington, IN USA},
Stephanie Debats\affil{1}{Princeton University, Princeton, NJ USA},
Tom Evans\affil{2}{Indiana University, Bloomington, IN USA},
Fanie Ferreira\affil{3}{GeoTerraImage, Pretoria, RSA},
Gabrielle Ragazzo\affil{1}{Princeton University, Princeton, NJ USA},
Justin Sheffield\affil{1}{Princeton University, Princeton, NJ USA}
\and
Kelly Caylor\affil{1}{Princeton University, Princeton, NJ USA}}

@@

\dropcap{T}he nature and distribution of landcover is a fundamental determinant of many environmental and social processes that drive or are affected by global change \cite{lambin_modelling_1997}, such as agricultural production and food security \cite{lark_cropland_2015,wright_recent_2013, licker_mind_2010}, carbon cycling \cite{asner_high-resolution_2010,gaveau_major_2014}, biodiversity loss \cite{newbold_global_2015,luoto_predicting_2004}, or demographic changes \cite{linard_assessing_2010}. Landcover maps are therefore critical for understanding the nature and impact of such changes \cite{see_improved_2015}, and they need to be accurate at the finest scales at which the underlying processes operate. For example, agricultural productivity and nutrient loadings can vary greatly between neighboring fields, and field sizes are often $<$2 hectares in regions where smallholder farming still dominates \cite{jain_mapping_2013, debats_generalized_????}. To understand agriculturally driven processes, it is thus necessary to accurately delineate fields at their smallest grain size, and to do so at regional to global scales to have a consistent set of maps.

Landcover data can only be developed with satellite imaging, but often the average size class of the cover type of interest is smaller than the sensor resolution, or spectrally indistinct from other neighboring covers, which propagates classification error \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. The result is that landcover datasets are generally inaccurate at finer scales and greatly differ between one another, particularly in those parts of the world undergoing the most rapid land use changes, where the aforementioned sources of bias tend to be most pronounced \cite{estes_projected_2013,fritz_comparison_2010,fritz_cropland_2011}.

These errors are well-known \cite{fritz_comparison_2010, fritz_cropland_2011, see_improved_2015, fritz_mapping_2015,verburg_challenges_2011}, and there are a variety of efforts underway to improve landcover maps, particularly for agriculture \cite{fritz_geo-wiki:_2012,estes_diylandcover:_2015}. What is less known is the degree to which these errors bias measurements built upon the distributional and areal information in landcover. An impediment to this understanding is that the errors are hard to quantify because spatially extensive reference data are not available for most regions of the world--particularly over Africa and other developing regions. Errors assessment therefore typically rely on a small number of ground truth points or survey data aggregated to political boundaries. For this reason, we have a better understanding of the biases between landcover datasets or in relation to country-level statistics \cite{fritz_comparison_2010,fritz_cropland_2011,kaptue_tchuente_comparison_2011} than we do of how error changes over spatial gradients or as a function of aggregation scale.

Being unable to fully quantify the errors in landcover maps of course makes it difficult, if not impossible, to quantify their impact on downstream analyses. There has been some work examining how such error influences climate simulations \cite{ge_impacts_2007}, agricultural land use patterns \cite{schmit_limitations_2006}, and carbon flux \cite{quaife_impact_2008} and human population estimates \cite{linard_assessing_2010}, but these either use simulated landcover errors \cite{ge_impacts_2007} or compare relevant differences in estimates between different satellite-derived landcover maps \cite{linard_assessing_2010, quaife_impact_2008}. The exception is \cite{schmit_limitations_2006}, who use a high quality, ground-collected reference map detailing farm land use parcels in central Belgium, but the number of sites and region were both fairly restricted, and the parcels were not spatially contiguous.

There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need in this study, using a unique, high accuracy agricultural landcover map for South Africa to quantify the errors in several latest generation landcover maps that are broadly used in global change studies.  We use this information to examine how i) landcover properties and related classification schemes influence error, ii) how these errors change with aggregation scale, with the specific goal of determining ``safe'' scales for making area-based calculations, and 3) how these errors propagate through several different forms of downstream analyses that broadly represent the global change research focus areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

\vspace{-0.5 cm}
\section{Study area and landcover data}
Our study focused on South Africa, which comprises nearly 6\% of sub-Saharan Africa's (SSA) landmass, and has a large, diverse agricultural sector, ranging from large commercial operations to smallholder farms \cite{hardy_rainfed_2011,estes_using_2014}. This diversity suggests that the country's agricultural landcover spans the range of types that are found throughout the rest of SSA.

The South African government commissioned a whole country cropland boundary map in order to stratifying the annual aerial crop type census used to calculate harvested area estimates \cite{fourie_better_2009}. The map was made by trained workers who visually interpreted high resolution satellite imagery and manually digitized field boundaries following a standardized mapping protocol. The resulting vectorized field maps, which were made in 2007 and updated in 2011, provide a unique, high accuracy reference dataset of both crop field distribution and size classes.  We converted the vector data into a rasterized estimated of cropland percentage at 1 km resolution (henceforth the ``reference map''), which was 97\% accurate in distinguishing cropped from non-cropped areas.

We compared our reference percent cropland estimates to those created from four satellite-derived landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite{fry_completion_2009},  MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013,shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and is the current state of the art for agricultural landcover maps. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates (hereafter simply the ``landcover maps''), resulting in 4 maps to compare to our reference cropland map (the ``reference map'').

\vspace{-0.5 cm}
\section{Quantifying Error}
We used these maps to first quantify error in cropland area estimates. We calculated error as the difference between the reference and landcover maps at different scales of aggregation (1 to 100 km), in order to estimate bias and how it varies with scale. Next, we assessed how bias correlates with the amount of cropland cover in agricultural landscapes, to gain insight into how landscape patterns may affect error.

We undertook five further analyses to investigate how map error can impact assessments founded on landcover maps. These include first-order analyses, in which values for a variable of interest are mapped to particular cover type(s), and second-order analyses, in which a process model draws on the cover types' values to calculate an output value. We created four datasets to represent first order analyses. The first was a series of maps of vegetated carbon stocks created following the methodology of Ruesch and Gibbs' \cite{ruesch_new_2008}. The second was constrained cropland percentage maps, which, following Ramankutty et al \cite{ramankutty_farming_2008} were adjusted so that their total cropland areas matched provincial-level reported cropland totals. Using these adjusted cropland percentage maps, we disaggregated district-reported maize harvested area and yields \cite[following][]{monfreda_farming_2008} . We then compared differences between total carbon stock estimates calculated from the reference map with those from the four cropland maps, and again examined how these differences changed as a function of aggregation scale.  We made the same comparisons for total maize harvested area, average yield, and total production.

For the second-order analyses, we examined how cropland cover errors influence 25 km resolution monthly evapotranspiration estimates produced using the Variable Infiltration Capacity \cite{liang_simple_1994} land surface hydrology model. For this example, we used the cropland maps to adjust the seasonally varying, landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these errors can impact the parameterization of an agent-based food security model \cite{chen_dependency_2013}. Spatially-explicit, agent-based models are frequently employed in land change science, and require an initialization step to assign landscape resources to model agents  \cite[e.g.][]{manson_agent-based_2007,evans_multi-scale_2004,kelley_relative_2011}. In this case, we used the cropland maps to allocate farmland to agents representing individual households in political districts, with the model assigning each household its initial cropland holdings using a function that considers total district cropland and how much cropland is near the agent's location.

@@

We created the 1 km reference and landcover maps, removing all croplands marked as communal or smallholder farmland in the reference vector maps (individual fields were not mapped with the same precision), as horticulture, or plantation forestry (SI), and calculated the total cropland extent in the remaining area (1,081,000 km$^2$, or 90\% of South Africa).  The 2011 reference map showed a cropland area of 104,310 km$^2$, which the SA-LC and GeoWiki maps overestimated by 26 and 5.8\%, respectively, and GlobCover and MODIS underestimated by 21 and 26.1\%.

We then aggregated each map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from the reference map at each scale of aggregation to asses the spatial patterns of bias (Fig. 1).

@@

The most pronounced extreme biases are in the MODIS and GlobCover maps, with both underestimating cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimating along the eastern to northern margins (red areas in Fig. 1). The average biases for GlobCover and MODIS are 21\% and 34\% at 1 km, meaning that each map underestimates cropland by that amount at this resolution. MODIS mean bias drops to 8\% at 50 km of aggregation, whereas GlobCover bias is still 24\% at 100 km (SI Appendix, Fig. S1).

The SA-LC map uniformly overestimates cropland throughout the country (Fig. 1), but the magnitude of mean bias is much lower at -8\% at 1 km to -6\% at 100 km  (Fig. S1). The GeoWiki map has a very heterogeneous pattern of bias, having a slight underestimation tendency (5\% mean bias) at 1 km, which changes to a smaller overestimation bias (-2 \%) at 100 km (Fig. S1).

\subsection{The role of landscape characteristics in shaping bias}
To investigate the degree to which landscape features influence landcover map bias, we extract all pixels in agricultural areas ($>$0.5\% cropland) of the 1 km reference map using the the boundaries of 354 magisterial districts (South Africa's finest administrative unit, which average 3,445 km$^2$ in size; SI Appendix, Fig. S3), and averaged the values of these pixels within each district. These provided an estimate of the intensity of cropping within the farmed parts of the landscape (with districts defining the landscape scale), as well as the degree of mixing between cropland and other land covers. We then extracted the cropland map biases for the same locations, and calculated the mean absolute bias within each district. Absolute bias describes the magnitude of bias but not its direction, and is more informative about the likelihood of the map being biased at any given point in the landscape than actual bias, as positive and negative biases can cancel each other out within small areas.

@@

\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between the mean absolute bias in cropland maps and cropland cover within agricultural areas (reference map pixels having $>$0.5\% cropland), averaged within the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different cropland maps, with the solid line indicating  predicted absolute bias, and the lighter shading the standard error of the coefficients. Models were also fit to the mean absolute biases across all four maps (black curve), and from all maps exlcuding GlobCover (grey surve).}
\label{afoto}
\end{figure}

\subsection{The impact of bias on calculating carbon stocks}
We used the data of \cite{ruesch_new_2008} to calculate carbon densities for African forests, shrublands, croplands, grasslands, and sparse habitats. We assigned the cropland carbon values to map cells in proportion to their cropland cover. For the non-cropland proportions, we assigned the carbon value from each of the other types, such that we created four  different carbon maps for each landcover map at each aggregation scale (Fig. S4), which allowed us to test how carbon estimates vary as a function of i) bias in the cropland estimates and ii) the characteristics of adjacent cover types.

At the 1 km scale, the differences between country-wide total carbon stocks made using any of the cropland maps and the reference were within +/-3\%, regardless of which cover type was adjacent to cropland (Table S1), because differences between total stock estimates are diluted by the large area ($\sim$50-70\%) of the country having no cropland. Comparing the carbon stocks from each map's agricultural area reveals much greater differences (Table S1). SA-LC overestimates carbon stocks by just 2\% when the adjacent cover type is forest, and up to 15\% when it is spare cover. MODIS ranges from negligible overestimates in denser carbon classes (forest, secondary forest, and shrublands) to 8-13\% underestimates for the grassland and sparse classes. GeoWiki underestimates for all types, from $<$1\% for sparse cover to 8\% for forest. GlobCover shows the largest total carbon stock bias, with over-estimates ranging from for sparse lands to 162\% for forest. The magnitude of this bias is due to false positives--GlobCover identified cropland in nearly 50\% of pixels, compared to 30\% for the other three cropland maps.

The spatial patterns of carbon bias (Fig. S4) are a reflection of the cropland biases (Fig. 1). Where cropland is underestimated and the surrounding cover type is of higher carbon density than cropland, carbon density is overestimated. For lower density (grassland and sparse vegetation), there may be slight underestimates.  These tendencies are reflected in the average of each map's spatial biases over agricultural areas (Fig. 3).  For example, MODIS and GlobCover overestimate carbon by $\sim$50\% on average at 1 km resolution when forest is the cropland-adjacent cover (stars in semi-transparent green and gold bars, Fig. 3; Table S2). When it is sparse vegetation (open circles in Fig. 3), MODIS underestimates by nearly 3\% at all scales, whereas maps that overestimate cropland (e.g. SA-LC, semi-transparent red) overestimate carbon density for this cover.  Overall, GeoWiki has the lowest average spatial biases, for all cover types and all resolutions. It's maximum bias is a 12\% overestimate at 1 km for forest cover, thereafter all biases are less than 10\%, and converge to +/-2\% above 1 km aggregation (Fig. 3, Table S2). All maps' biases are within +/-10\% bias after aggregation to 25 km.

\vspace{-0.9 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/carbon_veg_scale3.pdf}}
\caption{The average spatial biases in carbon density estimates based on cropland maps, expressed as a percent difference relative to the reference map. The values of actual bias (incorporating bias direction and magnitude) and absolute bias (magnitude only) are presented in semi-transparent and solid colors, respectively, with colors denoting the specific cropland map and symbols indicating which cover type was used to calculate cropland adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values are near zero and not shown for display clarity.}
\label{afoto}
\end{figure}

The patterns of mean absolute spatial bias (solid colored bars in Fig. 3) generally follow that of actual bias, but with somewhat higher magnitudes and a few important differences. The most notable is that Geo-Wiki has fairly large absolute bias at 1 km, averaging 27\% across across cover types (line in solid blue bar, Fig. 3, Table S2), which is close to the 36-37\% for GlobCover and MODIS. SA-LC has the lowest absolute bias across scales, averaging (across cover types) from 14\% at 1 km to 3\% at 100 km. (Fig. 3, Table S2). The increase in Geo-Wiki's absolute spatial bias relative to SA-LC's can be attributed to the spatially heterogeneous nature of its cropland biases (Fig. 1) compared to the other three cropland maps.

\subsection{Bias in harvested areas, yield, and production estimates}
We used the reference dataset to calculate total cropland area within South Africa's nine provinces, and then adjusted the the four cropland maps so that they each summed to the same area within each province \cite[per][]{ramankutty_farming_2008}. Despite this statistical constraint, the updated cropland maps still had substantial spatial biases similar in pattern (Fig. S5) to those in the unadjusted maps (Fig. 1), and we evaluated how these biases affected crop production estimates. We derived maize yield and harvested area estimates by disaggregating district-level  (n = 354, mean area = 3,445 km$^2$) agricultural census data \cite{statistics_south_africa_commercial_2007} following Monfreda \emph{et al.} \cite{monfreda_farming_2008}, and used that to calculate spatial biases in yield and production estimates at each aggregation scale.

The yields disaggregated onto the cropland maps were markedly different to those on the reference map, particularly in the lower density cropland areas in the center of the country, where GlobCover overestimated yields and MODIS and GeoWiki (to a lesser extent) underestimated them at 1 km resolution (Fig. S6). However, the mean spatial bias in yields at this resolution was pronounced only for GlobCover, equivalent to nearly 60\% of the mean reference yield of 3.4 tons ha$^{-1}$, compared to +/-5\% for the other maps (Fig. 5). Interestingly, GeoWiki and MODIS biases increased with aggregation, peaking at 10 km where both datasets underestimated yields by 30\%, thereafter declining to 10\% at 100 km, whereas GlobCover's mean yield bias declined linearly with aggregation (Fig. 5).

>>>>

%\usepackage{pnastwoF}
%\usepackage{cite}
\usepackage[numbers,round,sort,compress]{natbib}
\bibpunct{(}{)}{,}{n}{,}{,}  % https://xianblog.wordpress.com/tag/natbib/ (allows natbib with PNAS)
\usepackage{lineno}
\setlength\linenumbersep{2pt}
\usepackage{placeins}
\usepackage{float}

@@

\title{The Impact of Bias in Landcover Maps}
%\title{Our Understanding of Global Change is Built on a Shaky Foundation}

\author{Lyndon Estes\affil{1}{Princeton University, Princeton, NJ USA},
Peng Chen\affil{2}{Indiana University, Bloomington, IN USA},
Stephanie Debats\affil{1}{Princeton University, Princeton, NJ USA},
Tom Evans\affil{2}{Indiana University, Bloomington, IN USA},
Fanie Ferreira\affil{3}{GeoTerraImage, Pretoria, RSA},
Gabrielle Ragazzo\affil{1}{Princeton University, Princeton, NJ USA},
Justin Sheffield\affil{1}{Princeton University, Princeton, NJ USA}
Adam Wolf\affil{1}{Princeton University, Princeton, NJ USA}
\and
Kelly Caylor\affil{1}{Princeton University, Princeton, NJ USA}}

@@

\dropcap{T}he nature and distribution of landcover is an indicator that gives significant insight into socio-econonomic processes \cite{lambin_modelling_1997}, because human endeavors are so closely tied to how we transform land, whether it be the felling of ancient forests for farmland or erecting skycrapers. The vastness of our alteration of Earth’s landscapes suggests that landcover is a fundamental mediator of many environmental and social processes that drive or are affected by global change \cite{lambin_modelling_1997}, such as agricultural production and food security \cite{lark_cropland_2015,wright_recent_2013, licker_mind_2010}, carbon cycling \cite{asner_high-resolution_2010,gaveau_major_2014}, biodiversity loss \cite{newbold_global_2015,luoto_predicting_2004}, and changes in human demography \cite{linard_assessing_2010}. Like any view into nature, resolution and fidelity at fine scales are the keys to unlocking more granular and mechanistic insights into these processes \cite{see_improved_2015}. It is therefore unsurprising to see the explosive growth in private sector initiatives to develop new Earth observing capabilities, which range from small hobbyist drones\footnote{e.g. 3DRobotics, DJIA} to satellite arrays\footnote{Planet Labs, Skybox}, in order to add value to industries such as agriculture, mining, and construction. This rapid growth in fine-scale landcover mapping capability is creating new opportunities to develop actionable information for traditionally public-sector concerns, such as agricultural development\footnote{USAID's Feed the Future}, drought and flood adaptation\footnote{Global Index Insurance Facility, www.indexinsuranceforum.org}, and carbon cycle management\footnote{United Nations REDD+, www.un-redd.org/aboutredd}. But while the demand for more nuanced, landcover-based insights is growing, there is only now the opportunity to use finer-scaled imagery to comprehensively interrogate the accuracy and biases in the landcover products that have become ubiquitous in global change research.

Global landcover data can only practically be derived from satellite imaging, but in many regions the average size class of the cover type of interest is smaller than the sensor resolution, or spectrally indistinct from other neighboring covers, which propagates classification error \cite{see_improved_2015,lobell_use_2013,estes_diylandcover:_2015}. The result is that landcover maps are generally inaccurate at finer scales and disagree greatly, particularly in those parts of the world undergoing the most rapid land use changes, where these sources of bias tend to be most pronounced \cite{estes_projected_2013,fritz_comparison_2010,fritz_cropland_2011}.

Errors in landcover products are widely-acknowledged \cite{fritz_comparison_2010, fritz_cropland_2011, see_improved_2015, fritz_mapping_2015,verburg_challenges_2011}, and there are a variety of efforts underway to improve landcover maps, particularly for agriculture \cite{fritz_geo-wiki:_2012,estes_diylandcover:_2015}. What is less known is the degree to which these errors bias analyses derived from the distributional and areal information in landcover. Errors are hard to quantify because spatially extensive reference data are not available for most regions of the world--particularly over Africa and other developing regions. Error assessments therefore typically rely on a small number of ground truth points for a bottom-up assessment or aggregated survey data for a top-down sanity check. For this reason, we have a better understanding of discrepancies between landcover datasets in relation to country-level statistics \cite{fritz_comparison_2010,fritz_cropland_2011,kaptue_tchuente_comparison_2011}, which offers little direction for how to arrive at a true number.

Being unable to fully quantify the errors in landcover maps of course makes it difficult, if not impossible, to quantify their impact on downstream analyses. There has been some work examining how such error influences climate simulations \cite{ge_impacts_2007}, agricultural land use patterns \cite{schmit_limitations_2006}, and carbon flux \cite{quaife_impact_2008} and human population estimates \cite{linard_assessing_2010}, but these either use simulated landcover errors \cite{ge_impacts_2007} or compare relevant differences in estimates between different satellite-derived landcover maps \cite{linard_assessing_2010, quaife_impact_2008}. One exception is a study  \cite{schmit_limitations_2006} that used a high quality, ground-collected reference map detailing farm land use parcels in central Belgium, but the number of sites and region were both fairly restricted, and the parcels were not spatially contiguous.

Just as a tall edifice cannot be built on a shaky foundation, global change science needs to be based on sound landcover data. There is thus an urgent need to more precisely quantify landcover map errors and how these vary over large regions, particularly for the regions where landcover is changing most rapidly yet is most poorly known.  We address this need in this study, using a unique, high accuracy agricultural landcover map for the entire country of South Africa to quantify the errors in several latest generation landcover maps that are widely used in global change studies.  We use this information to examine how i) landcover properties and related classification schemes influence error, ii) how these errors change with aggregation scale, with the specific goal of determining ``safe'' scales for drawing area-based inferences, and iii) how these errors propagate through several different forms of downstream analyses that broadly represent the global change research focus areas, including biogeochemical and land use change studies, food security assessments, land surface hydrology and climatology, and human geography.

\vspace{-0.5 cm}
\section{Study area and landcover data}
South Africa comprises nearly 6\% of sub-Saharan Africa's (SSA) landmass, and has a large, diverse agricultural sector, ranging from large commercial operations to smallholder farms \cite{hardy_rainfed_2011,estes_using_2014}. This diversity suggests that the country's agricultural landcover spans the range of types that are found throughout the rest of SSA.

The South African government commissioned a whole country cropland boundary map to enhance its annual collection of agricultural statistics \cite{fourie_better_2009}. The map was made by trained workers who visually interpreted high resolution satellite imagery ($<$5 m SPOT imagery) and manually digitized field boundaries following a standardized mapping protocol. The resulting vectorized field maps, which were made in 2007 and updated in 2011, provide a unique, high quality reference dataset describing crop field distributions and size classes, and are 97\% accurate in distinguishing cropland from non-cropland at 200 m resolution. We intersected the field vectors with a 1 km grid, and calculated the percent of each cell occupied by fields to create a gridded cropland reference map.

We compared the reference map with similar maps derived from four existing landcover datasets. We obtained South Africa's 30 m resolution National Landcover map (SA-LC) for 2009 \cite{sanbi_national_2009}, the 500 m resolution MODIS Landcover for 2011 \cite{land_processes_distributed_active_archive_center_lp_daac_modis_2011, friedl_modis_2010}, the 300 m resolution GlobCover 2009 \cite{arino_global_2012}, and the new 1 km Geo-wiki hybrid-fusion cropland map for Africa \cite{fritz_mapping_2015}. We chose these particular datasets because they are nearly contemporaneous with our reference data, and represent the major types of landcover products used by researchers: SA-LC typifies the higher resolution, Landsat-derived maps that are developed individually for many countries \cite[e.g.][]{fry_completion_2009}, MODIS and GlobCover are widely used global-scale products \cite{gross_monitoring_2013,shackelford_conservation_2015}, while Geo-Wiki incorporates the first three datasets and represents the current state-of-the-art in landcover mapping. We extracted the cropland classes from the first three datasets and converted these to 1 km resolution percent cropland estimates (hereafter simply the ``cropland maps''), resulting in 4 maps to compare to our reference map.

\vspace{-0.5 cm}
\section{Quantifying Error and Bias}
We first quantified the errors in cropland area estimates based on the pixel-wise differences between the reference map and each of the four cropland maps. We calculated these errors for five different scales of aggregation, from the original 1 km up to 100 km, in order to calculate bias--the average pixel-wise error--and how it varies with scale. Next, we assessed the degree to which the pattern of agricultural landscapes, expressed in terms of the mean cropland cover, impacts bias.

We then investigated how these landcover map biases propagate through data that are built upon them. We began by examining the biases in three different kinds of landcover-based data that we consider to be ``first-order'', in that they are derived by mapping a variable of interest to a landcover type(s). The first of these was to assess potential biases in Tier-1 carbon accounting maps that we constructed from the cropland maps, following \cite{ruesch_new_2008}. The second was an assessment of gridded maize yields derived by disaggregating district-scale agricultural census data for both maize yield and harvested area \cite[following][]{monfreda_farming_2008,ramankutty_farming_2008}. We used these same data to calculate gridded maize production estimates, which we used to conduct the third first order analysis.

Finally, we conducted two second-order analyses, in which a process model draws on the cover types' values to calculate an output value. For the first of these, we examined how cropland map biases influence 25 km resolution monthly evapotranspiration calculated by the Variable Infiltration Capacity \cite{liang_simple_1994} land surface hydrology model. For this example, we used the cropland maps to adjust landcover-specific leaf area index (LAI) values that VIC uses to partition water vapor fluxes into their evaporative and transpirative components. In the second example, we examined how these errors can impact the parameterization of an agent-based food security model \cite{chen_dependency_2013}. Spatially-explicit, agent-based models are frequently employed in land change science, and require an initialization step to assign landscape resources to model agents  \cite[e.g.][]{manson_agent-based_2007,evans_multi-scale_2004,kelley_relative_2011}. In this case, we used cropland maps to provide the model the location and abundance of cropland, which is used to  allocate an initial share of cropland to each farm household (agent) in the model.

@@

We created the 1 km reference map after removing all field types classified as communal/smallholder agriculture (individual fields in this category were not mapped, thus they were removed to prevent potential cropland area overestimates) or permanent tree-crops (SI), and calculated the total cropland extent in the remaining area (1,081,000 km$^2$, or 90\% of South Africa). The 2011 reference map showed a cropland area of 104,310 km$^2$, which the SA-LC and GeoWiki maps overestimated by 26 and 5.8\%, respectively, and GlobCover and MODIS underestimated by 21 and 26.1\%.

We then aggregated the reference and each cropland map to 5, 10, 25, 50, and 100 km resolutions, and subtracted the four landcover maps from the reference map at each scale of aggregation to assess patterns of differences (Fig. 1). Negative pixels here represent overestimation error, while positive values indicate underestimates.

@@

The most pronounced differences were in the MODIS and GlobCover maps, which both underestimated cropland extent by 10-75\% in the center of the country (blue areas in Fig. 1, and the dominant production region), and overestimated along the eastern to northern margins (red areas in Fig. 1). The differences between the reference and these two maps at 1 km resolution were respectively 21\% (MODIS) and 34\% (GlobCover), on average. This mean pixel-wise error expresses the amount of \emph{bias} in a map, that is its tendency to err in a particular direction. For MODIS, bias drops to 8\% at 50 km of aggregation, whereas GlobCover bias is 24\% at 100 km (SI Appendix, Fig. S1).

The SA-LC map uniformly overestimated cropland throughout the country (Fig. 1), but was less biased, ranging from -8\% at 1 km to -6\% at 100 km  (Fig. S1). The GeoWiki map has a strongly heterogenous pattern of differences (Fig. 1) and the smallest magnitude of bias, which changed between slight tendencies to underestimate (5\% at 1 km) and overestimate (-2 \% at 100 km,  Fig. S1).

\subsection{The role of landscape characteristics in shaping bias}
To investigate the degree to which landscape features influence landcover map bias, we extracted all pixels in agricultural areas ($>$0\% cropland) of the 1 km reference map using the boundaries of 354 magisterial districts (South Africa's finest administrative unit, which average 3,445 km$^2$ in size; SI Appendix, Fig. S3), and calculated the district-wise mean for these pixels to provide a measure of cropland density, and information on the degree of mixing between cropland and other land covers. We then extracted the cropland map errors for the same pixels, and calculated the absolute value of bias for each district. Absolute bias describes the magnitude of bias but not its direction, and is more informative about the likelihood of the map being biased at any given point in the landscape than actual bias, as positive and negative biases can cancel each other out within small areas.

@@

%\vspace{-1 cm}
\begin{figure}[ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/biases_md_lnorm_gam_mu0.pdf}}
\caption{The relationship between the mean absolute bias in cropland maps and cropland cover within agricultural areas (reference map pixels having $>$0.5\% cropland), averaged within the boundaries of magisterial districts (n = 345), as fit with a generalized additive model. Prediction curves are color-coded to the different cropland maps, with the solid line indicating  predicted absolute bias, and the lighter shading the standard error of the coefficients. Models were also fit to the mean absolute biases across all four maps (black curve), and from all maps exlcuding GlobCover (grey surve).}
\label{afoto}
\end{figure}

\subsection{The impact of bias on calculating carbon stocks}
We used the data of \cite{ruesch_new_2008} to calculate carbon densities for African forests, shrublands, croplands, grasslands, and sparse habitats (semi-arid grasslands and low shrublands), assigning cropland carbon values to map cells in proportion to their cropland cover. For the non-cropland proportions, we assigned the carbon value from each of the other types, such that we created four different carbon maps for each landcover map at each aggregation scale (Fig. S4), which allowed us to test how carbon estimates vary as a function of i) cropland map bias and ii) the characteristics of adjacent cover types.

The difference between total carbon stocks for the country made using any of the cropland maps were within +/-3\% of those based on the reference map, regardless of which cover type was adjacent to cropland (Table S1), because the large of non-cropland in the country ($\sim$50-70\%, Fig. 1) dilutes any map errors. Comparing total stocks between maps for just the agricultural area (30-50\% of the country) reveals much greater differences (Table S1). SA-LC overestimated carbon stocks by just 2\% when the adjacent cover type was forest, and up to 15\% when it was sparse cover. MODIS ranged from negligible differences in denser carbon classes (forest, secondary forest, and shrublands) to 8-13\% underestimates for the grassland and sparse classes. GeoWiki underestimated for all types, from $<$1\% for sparse cover to 8\% for forest. GlobCover grossly overestimated total carbon stocks for agricultural areas, varying from 64\% for sparse lands to 162\% for forest. The magnitude of this bias was due to false positives--GlobCover identified cropland in nearly 50\% of pixels, compared to 30\% for the other three cropland maps.

The spatial patterns of errors in carbon estimates (Fig. S4) reflect those of cropland biases (Fig. 1). Where cropland was underestimated and the surrounding cover type was of higher carbon density than cropland, carbon density was overestimated. For lower density cover (grassland and sparse vegetation), carbon stocks were underestimated, but by small magnitudes. These tendencies were reflected in each map's biases, calculated over agricultural areas (Fig. 3). For example, MODIS and GlobCover bias was $\sim$-50\% (overestimation) at 1 km resolution when forest was the cropland-adjacent cover (stars in semi-transparent green and gold bars, Fig. 3; Table S2). For sparse vegetation (open circles in Fig. 3), MODIS bias was 3\% at all scales, whereas maps that overestimated cropland (e.g. SA-LC, semi-transparent red) overestimated carbon density for this cover type, because cropland has a higher carbon density \cite{ruesch_new_2008}. Overall, GeoWiki had the lowest bias, for all cover types and all resolutions. It worst bias was a tendency to overestimate by 12\% at 1 km when forest was adjacent, but at coarser scales this bias reduced to just a few percent (Fig. 3, Table S2). All maps' biases are within +/-10\% bias after aggregation to 25 km.

%\vspace{-0.9 cm}
\begin{figure}[!ht]
\centerline{\includegraphics[width=.4\textwidth]{figures/carbon_veg_scale3.pdf}}
\caption{The biases in carbon density estimates based on cropland maps, expressed as a percent difference relative to the reference map. The values of actual bias (incorporating bias direction and magnitude) and absolute bias (magnitude only) are presented in semi-transparent and solid colors, respectively, with colors denoting the specific cropland map and symbols indicating which cover type was used to calculate cropland adjacent carbon density. The bar represents the mean biases calculated across each of the 5 cover types. Shrubland and grassland bias values are near zero and not shown for display clarity.}
\label{afoto}
\end{figure}

Absolute bias in carbon maps (solid colored bars in Fig. 3) generally followed the same patterns, but with somewhat higher magnitudes and a few important differences. The most notable is that Geo-Wiki had fairly large absolute bias at 1 km, averaging 27\% across across cover types (line in solid blue bar, Fig. 3, Table S2), which is close to the 36-37\% for GlobCover and MODIS. SA-LC had the lowest absolute bias across scales, averaging (across cover types) from 14\% at 1 km to 3\% at 100 km. (Fig. 3, Table S2). The increase in Geo-Wiki's absolute bias relative to SA-LC's can be attributed to the highly heterogeneous nature of its cropland errors (Fig. 1), which varied between positive and negative over much shorter distances than the other three cropland maps.

\subsection{Bias in harvested areas, yield, and production estimates}
The disaggregated yield and harvested area maps of \cite{monfreda_farming_2008} are built upon cropland fraction maps where the total area is adjusted to match survey-derived agricultural area statistics reported for administrative districts \cite[provinces, in South Africa's case][]{ramankutty_farming_2008}. To be consistent with this methodology, we adjusted our cropland maps using the same procedure, but we used the reference map to calculate total cropland area for each of South Africa's nine provinces, and then updated the pixel-wise cropland percentages in the four cropland maps so that the province-wise sums matched the reference areas \cite[][and see SI]{ramankutty_farming_2008}. Despite this statistical constraint, the updated cropland maps still had large spatial errors that were similar in pattern (Fig. S5) to those in the unadjusted maps (Fig. 1), and we evaluated how these residuals affected gridded estimates of the yield and production of maize, South Africa's largest crop \cite{estes_comparing_2013}. To create these maps, we followed \cite{monfreda_farming_2008} by disaggregating district-level (n = 354, mean area = 3,445 km$^2$) agricultural census data \cite{statistics_south_africa_commercial_2007} for maize yield and harvested area, aggregated each set of maps, and multiplied the two to calculate production at each scale.
